\chapter{dataset objects} \label{chap_data}
\label{datasetobj} \index{Dataset objects} \index{Dataset}


{\em Dataset objects} are used to store, manage, and manipulate
data. A new {\em dataset object} is created by typing

#> dataset# {\em objectname}

where {\em objectname} is the name to be assigned to the data set.
After the creation of a {\em dataset object} you can apply the
methods discussed below.

Note that in the current version of {\em BayesX} only numerical
variables are allowed. String valued variables, for example, are
not yet supported by {\em BayesX} and the attempt to read such
variables will raise an error message.

\clearpage

\section{Method descriptive} \label{descriptive} \index{Summary
statistics} \index{Descriptives} \index{Dataset!Descriptive command}

\begin{stanza}{Description}

Method #descriptive# calculates and displays univariate summary
statistics. To be more specific, the method computes the number of
observations, the mean, median, standard deviation, minimum and
maximum of the specified variables.
\end{stanza}

\begin{stanza}{Syntax}

 #> #{\em objectname}#.descriptive# {\em varlist} [#if# {\em expression}]

Method #descriptive# computes summary statistics for the variables
in {\em varlist}. An optional #if# statement may be added to
analyze only a part of the data.
\end{stanza}

\begin{stanza}{Options}

not allowed
\end{stanza}


\begin{stanza}{Example}

{The statement

#> d.descriptive x y#

computes summary statistics for the variables #x# and #y#.
The statement

#> d.descriptive x y if x>0#

restricts the analysis to observations with #x>0#.}
\end{stanza}

\clearpage

\section{Method drop}
\label{drop} \index{Dropping variables} \index{Dropping
observations} \index{Dataset!Drop command}

\begin{stanza}{Description}

Method #drop# deletes variables or observations from the data set.
\end{stanza}

\begin{stanza}{Syntax}

#> #{\em objectname}.#drop# {\em varlist}

#> #{\em objectname}.#drop if# {\em expression}

The first command is used to delete the variables specified in
{\em varlist} from the data set. The second statement deletes
certain observations, i.e. all observations for which {\em
expression} is true will be deleted.
\end{stanza}


\begin{stanza}{Options}

{not allowed}
\end{stanza}


\begin{stanza}{Example}

The statement

#> credit.drop account duration#

deletes the variables #account# and #duration# from the credit
scoring data set. With the statement

#> credit.drop if marstat = 2#

all observations with #marstat = 2#, i.e. all persons living
alone, will be deleted. The statement

#> credit.drop account duration if marstat = 2#

will raise the error

#ERROR: dropping variables and observations in one step not allowed#

because is not allowed to drop variables and observations in one
single command.
\end{stanza}

\clearpage

\section{Functions and Expressions}
\label{expression} \index{Expressions}

The primary use of expressions is to generate new variables or
change existing variables, see \autoref{generate} and
\autoref{replace}, respectively. Expressions may also be used in
#if# statements to force {\em BayesX} to apply a method only to
observations where the boolean expression in the #if# statement is
true. The following statements are all examples of expressions:

#2+2# \\
#log(amount)# \\
#1*(age <= 30)+2*(age > 30 & age <= 40)+3*(age > 40)# \\
#age=30# \\
#age+3.4*age^2+2*age^3# \\
#amount/1000#


\subsection{Operators}
\index{Expressions!Operators} \index{Operators}

Three different types of operators can be included in expressions:
arithmetic, relational and logical operators.

\subsubsection{Arithmetic operators}
\index{Operators!Arithmetic}

The arithmetic operators are #+# (addition), #-# (subtraction),
#*# (multiplication), #/# (division), #^# (raise to a power) and
the prefix #-# (negation). Any arithmetic operation on a missing
value or an undefined arithmetic operation (such as division by
zero) yields a missing value.

\begin{stanza}{Example}

The expression

#(x+y^(3-x))/(x*y)#

denotes the formula
$$
\frac{x+y^{3-x}}{x\cdot y}
$$
and evaluates to missing if #x# or #y# is missing or zero.
\end{stanza}

\subsubsection{Relational operators}
\index{Operators!Relational}

The relational operators are #># (greater than), #<# (less than),
#>=# (greater than or equal), #<=# (less than or equal), #=#
(equal) and #!=# (not equal). Relational expressions are either 1
(if the expression is true) or 0 (if the expression is false).

\begin{stanza}{Example}

Relational operators can for example be used to create indicator
variables. The following statement generates a new variable
#amountcat#, with value 1 if #amount<=10# and value 2 if
#amount>10#.

#> credit.generate amountcat = 1*(amount<=10)+2*(amount>10)#

Another useful application of relational operators is in #if#
statements. For example, changing an existing variable only when a
certain condition holds can be achieved by the following command:

#> credit.replace amount = NA if amount <= 0#

This sets all observations missing where #amount<=0#.
\end{stanza}

\subsubsection{Logical operators}
\index{Operators!Logical}

The logical operators are #&# (and) and #|# (or).

\subsubsection*{Example}

Suppose you want to generate a variable #amountind# whose value is
1 for married people with amount greater than 10 and 0 otherwise.
This can be achieved by typing

#> credit.generate amountind = 1*(marstat=1 & amount>10)#

\subsubsection{Order of evaluation of the operators}
\index{Operators!Order of evaluation}

The order of evaluation (from first to last) of operators is

#^# \\
#/#, #*#\\
#-#, #+#\\
#!=#, #>#, #<#, #<=#, #>=#, #=#\\
#&#, #|#

Brackets can be used to change the order of evaluation.


\subsection{Functions}
\index{Functions}

Functions are a further component of expressions. Any function is
indicated by the function name followed by an opening and a
closing parenthesis. Inside the parentheses one or more arguments
may be specified. The argument(s) of a function may again be
expressions, including calls to further functions. Multiple
arguments of a function are separated by commas. All functions
return missing values when given missing values as arguments or
when the result is undefined.

All mathematical function currently available in {\em BayesX} are
referenced in \autoref{mathfunc}. Statistical functions can be
found in \autoref{statfunc}.

\index{Functions!abs} \index{Functions!cos} \index{Functions!sin}
\index{Functions!exp} \index{Functions!floor} \index{Functions!lag}
\index{Functions!logarithm} \index{Functions!square root}
\index{Functions!Bernoulli distributed random numbers}
\index{Functions!Binomial distributed random numbers}
\index{Functions!Exponential distributed random numbers}
\index{Functions!Normally distributed random numbers}
\index{Functions!Uniformly distributed random numbers}

\begin{table}[ht]
\begin{center}
\begin{tabular}{|l|l|}
\hline
{\bf Function} & {\bf Description} \\
\hline \hline
#abs(#x#)# & absolute value \\
#cos(#x#)# & cosine of radians \\
#exp(#x#)# & exponential \\
#floor(#x#)# & returns the integer obtained by truncating $x$. \\
& Thus #floor(5.2)# evaluates to 5 as does #floor(5.8)#. \\
#lag(#x#)# & lag operator \\
#log(#x#)# & natural logarithm \\
#log10(#x#)# & log base 10 of $x$ \\
#sin(#x#)# & sine of radians \\
#sqrt(#x#)# & square root \\
\hline
\end{tabular}
{\em\caption{\label{mathfunc} List of mathematical functions.}}
\end{center}
\end{table}



\begin{table}[ht]
\begin{center}
\begin{tabular}{|l|p{11cm}|}
\hline
{\bf Function} & {\bf Description} \\
 \hline \hline
 #bernoulli(#$p$#)# & returns Bernoulli distributed
 random numbers with probability of success $p$. If $p$ is not
 within the interval $[0;1]$, a
 missing value will be returned. \\
 \hline
 #binomial(#$n$#,#$p$#)# & returns $B(n;p)$ distributed random
 numbers. Both, the number of trials $n$ and the probability of
 success $p$ may be expressions. If $n < 1$, a missing value will
 be returned. If $n$ is not integer valued, the number of trials
 will be $[n]$. If $p$ is not within the interval $[0;1]$, a
 missing value will be returned. \\
 \hline
 #exponential(#$\lambda$#)# & returns exponentially distributed
 random numbers with parameter $\lambda$.
 If $\lambda \leq 0$, a missing value will be returned. \\
 \hline
 #normal()# & returns standard normally distributed random
 numbers;
 $N(\mu,\sigma^2)$ distributed random numbers may be generated with $\mu + \sigma$*normal(). \\
 \hline
 #uniform()# & uniform
 pseudo random number function; returns uniformly distributed
 pseudo-random numbers on the interval $(0,1)$ \\
 \hline
\end{tabular}
{\em \caption{\label{statfunc} List of statistical functions}}
\end{center}
\end{table}


\subsection{Constants}
\index{Expressions!Constants} \index{Number of observations}
\index{Missing values} \index{$\pi$} \index{Current observation}

\autoref{constant} lists all constants that can be used in
expressions.


\begin{table}[ht]
\begin{center}
\begin{tabular}{|l|l|}
\hline
Constant & Description \\
\hline \hline
\texttt{\_n} & contains the number of the current observation.  \\
\texttt{\_N }& contains the total number of observations in the data set. \\
\texttt{\_pi} & contains the value of $\pi$. \\
\texttt{NA} & indicates a missing value \\
.  & indicates a missing value \\

\hline
\end{tabular}
{\em \caption{\label{constant} List of constants}}
\end{center}
\end{table}


\subsubsection*{Example}

The following statement generates a variable #obsnr# whose value
is 1 for the first observation, 2 for the second and so on.

#> credit.generate obsnr = _n#

The command

#> credit.generate nrobs = _N#

generates a new variable {\em nrobs} whose values are all equal to
the total number of observations, say 1000, for all observations.

\subsection{Explicit subscribing}
\index{Expressions!Explicit subscribing} \index{Subscribing}

Individual observations on variables can be referenced by
subscribing the variables. Explicit subscripts are specified by
the variable name with square brackets that contain an expression.
The result of the subscript expression is truncated to an integer,
and the value of the variable for the indicated observation is
returned. If  the value of the subscript expression is less than 1
or greater than the number of observations in the data set,
a missing value is returned.

\subsubsection*{Example}

Explicit subscribing combined with the variable #_n# (see
\autoref{constant}) can be used to create lagged values of a
variable. For example, the lagged value of a variable #x# in a
data set #data# can be created by

#> data.generate xlag = x[_n-1]#

Note that #xlag# can also be generated using the #lag# function

#> data.generate xlag = lag(x)#


\clearpage

\section{Method generate}
\label{generate} \index{Generating new variables}
\index{Dataset!Generate command}

\begin{stanza}{Description}

Method #generate# is used to create a new variable in an existing
{\em dataset object}.
\end{stanza}


\begin{stanza}{Syntax}

#> #{\em objectname}.#generate# {\em newvar} = {\em expression}

Method #generate# creates a new variable with name {\em newvar}
(compare \autoref{varnames} for valid variable names) The values
of the new variable are specified by {\em expression}.
\end{stanza}


\begin{stanza}{Options}

not allowed
\end{stanza}


\begin{stanza}{Example}

The following command generates a new variable called #amount2#
which contains the squared amount in the credit scoring data set.

#> credit.generate amount2 = amount^2#

If you try to change the variable currently generated, for example by typing

#> credit.generate amount2 = amount^0.5#

the error message

#ERROR: variable amount2 is already existing #

will occur. This prevents you to change an existing variable
unintentionally. An existing variable can be changed with method
#replace#, see \autoref{replace}.

If you want to generate an indicator variable #largeamount# for
amounts exceeding a certain value, say 3.5, you have to type

#> credit.generate largeamount = 1*(amount>3.5)#

The variable #largeamount# takes the value 1 if #amount# is larger
than 3.5 and 0 otherwise.

\end{stanza}

\clearpage

\section{Method infile}
\label{infile} \index{Reading data from ASCII files}
\index{Dataset!Infile command}

\begin{stanza}{Description}

Reads data saved in an ASCII file.
\end{stanza}

\begin{stanza}{Syntax}

#> #{\em objectname}.#infile #[{\em varlist}] [{\em , options}]
#using# {\em filename}

Reads the data stored in {\em filename}. The variable names can
either be specified in {\em varlist} or be extracted from the
first line of the file. To be more specific, if {\em varlist} is
empty it is assumed that the first row of the file contains the
variable names separated by blanks or tabs. The observations do
not have to be stored in a special format, except that successive
observations should be separated by one or more blanks or tabs.
The first value read from the file will be the value of the first
variable of the the first observation, the second value will be
the value of the second variable of the first observation, and so
on. An error message will occur if no values can be read for the
last observation for some variables.

The period '.' or '#NA#' should be used to indicate missing
values.

Note that in the current version of {\em BayesX} only numerical
variables are allowed. Thus, the attempt to read string valued
variables, for example, will cause an error.
\end{stanza}

\subheader{Options}

\begin{itemize}
\item #missing = # {\em missingsigns} \\
By default the period '.' or '#NA#' indicate missing values. If
the missing values are indicated by a different sign, you can
specify them in the #missing# option. For example #missing = MIS#
defines '#MIS#' as an indicator for a missing value. Note that
 '.' and '#NA#' remain valid indicators for missing values, even if the missing
option is specified.

\item #maxobs = #{\em integer} \\
If you work with large data sets, you may observe the problem that
reading in a data set using the #infile# command is very time
consuming. The reason for this problem is that {\em BayesX} does
not know the number of observations in advance and therefore is
unable to allocate enough memory. Therefore new memory has to be
allocated whenever a certain amount of memory is exceeded.
Specifying option #maxobs# allows to circumvent this problem and
to considerably reduce the computing time. When #maxobs# is
specified, {\em BayesX} can allocate enough memory to store at
least {\em integer} observations before new memory must be
reallocated. Suppose for example that your data set consists of
approximately 100,000 observations. Then specifying
#maxobs=105000# allocates enough memory to read in the data set
quickly. Note that #maxobs=105000# does not require the data set
to have exactly 105,000 observations. It only means that new
memory will have to be allocated when the number of observations
exceeds the 105,000 observations limit.
\end{itemize}



\begin{stanza}{Example}

Suppose we want to read a data set stored in
#c:\data\testdata.raw# containing the two variables #var1# and
#var2#. The first few rows of the datafile could look like this:

var1 var2 \\
2 2.3 \\
3 4.5 \\
4 6 \\
...


To read the data set, we have to create a new {\em dataset object}
first, say #testdata#. We proceed by reading the data using the
#infile# command:

#> dataset testdata# \\
#> testdata.infile using c:\data\testdata.raw#

If the first row of the data set file contains no variable names,
the second command has to be altered to:

#> testdata.infile var1 var2 using c:\data\testdata.raw#

If furthermore the data set is large with about 100,000
observations. In this case, the #maxobs# option is very useful to
reduce reading time. Typing for example

#> testdata.infile var1 var2 , maxobs=101000 using c:\data\testdata.raw#

will be much faster than the command without option #maxobs#.
\end{stanza}

\clearpage



\section{Method outfile}
\label{outfile} \index{Writing data to a file} \index{Saving data in
an ASCII file} \index{Dataset!Outfile command}

\begin{stanza}{Description}

Method #outfile# writes data to a file in ASCII format.
\end{stanza}



\begin{stanza}{Syntax}

{#> #{\em objectname}.#outfile# [{\em varlist}] [#if# {\em expression}] [{\em , options}] #using# {\em filename}

#outfile# writes the variables specified in {\em varlist} to the
file with name {\em filename}. If {\em varlist} is omitted in the
outfile statement, {\em all} variables in the data set will be
included. Each row in the data file will correspond to one
observation. Different variables will be separated by blanks.
Optionally, an #if# statement may be used to store only those
observations where a certain boolean expression is true.}
\end{stanza}


\subheader{Options}


\begin{itemize}
\item #header# \\
Specifying the #header# option causes {\em BayesX} to write the
variable names to the first row of the created data file.
\item #replace# \\
The #replace# option allows {\em BayesX} to overwrite an already
existing data file. If #replace# is omitted and the file specified
in {\em filename} is already existing, an error will be raised.
This prevents you from unintentionally overwriting an existing
file.
\end{itemize}


\begin{stanza}{Example}

The statement

#> credit.outfile using c:\data\cr.dat#

writes the complete credit scoring data set to
#c:\data\cr.dat#. To generate two different
ASCII
data sets for married people and people living alone, you could type

#> credit.outfile if marstat = 1 using c:\data\crmarried.dat# \\
#> credit.outfile if marstat = 2 using c:\data\cralone.dat#

If you want to store only the two variables #y# and #amount#, you
could type

#> credit.outfile y amount using c:\data\cr.dat#

This will raise the error message

#ERROR: file c:\data\cr.dat is already existing#

because #c:\data\cr.dat# has already been created. You can
overwrite the file using the #replace# option

#> credit.outfile y amount, replace using c:\data\cr.dat#
\end{stanza}



\clearpage




\section{Method rename}
\index{Renaming variables} \index{Dataset!Rename command}


\label{rename}


\begin{stanza}{Description}

#rename# is used to change variable names.
\end{stanza}


\begin{stanza}{Syntax}

#> #{\em objectname}.#rename# {\em varname} {\em newname}

#rename# changes the name of {\em varname} to {\em newname}. {\em
newname} must be a valid variable name, see \autoref{varnames} for
valid variable names.
\end{stanza}

\begin{stanza}{Options}

not allowed

\end{stanza}

\clearpage



\section{Method replace}
\label{replace} \index{Changing existing variables}
\index{Dataset!Replace command}



\begin{stanza}{Description}

{#replace# changes values of an existing variable.}
\end{stanza}


\begin{stanza}{Syntax}

 #> #{\em objectname}.#replace# {\em varname} = {\em expression} [#if# {\em expression}]

#replace# changes some or all values of the variable {\em varname}
to the values specified in {\em expression}. If {\em varname} is
not existing, an error will be raised. An optional #if# statement
may be used to change the values of the variable only if the
boolean expression {\em expression} is true.
\end{stanza}


\begin{stanza}{Options}

not allowed
\end{stanza}


\begin{stanza}{Example}

The statement

#> credit.replace amount = NA if amount<0#

changes the values of the variable #amount# in the credit scoring
data set to missing if #amount<0#.
\end{stanza}



\clearpage



\section{Method set obs}
\label{setobs}


\begin{stanza}{Description}

{#set obs# changes the number of observations in a data set.}
\end{stanza}


\begin{stanza}{Syntax}

#> #{\em objectname}.#set obs# = {\em intvalue}

#set obs# raises the number of observations in the data set to
{\em intvalue}. The new number of observations has to be greater
or equal than the current number to prevent you from
unintentionally deleting parts of the data currently in memory.
Observations may be eliminated using the #drop# statement, see
\autoref{drop}. The values of newly created observations will be
set to the missing value.
\end{stanza}



\clearpage



\section{Method tabulate}
\label{tabulate} \index{Tabulate} \index{Table of frequencies}
\index{One way table of frequencies} \index{Dataset!Tabulate
command}

\begin{stanza}{Description}

Method #tabulate# calculates and displays a frequency tables.
\end{stanza}

\begin{stanza}{Syntax}

#># {\em objectname}.#tabulate# {\em varlist} [#if# {\em
expression}]

Method #tabulate# computes and displays frequency tables of the
variables specified in {\em varlist}. An optional #if# statement
may be added to restrict the analysis to a part of the data.
\end{stanza}

\begin{stanza}{Options}

not allowed
\end{stanza}

\begin{stanza}{Example}

The statement

#> d.tabulate x y#

displays frequency tables for the variables #x# and #y#. The
statement

#> d.tabulate x y if x>0#

restricts the analysis to observations with #x>0#.
\end{stanza}



\clearpage



\section{Variable names}
\label{varnames} \index{Variables names}


A valid variable name is a sequence of letters (#A#-#Z# and
#a#-#z#), digits (#0#-#9#), and underscores (#_#). The first
character of a variable name must be either a letter or an
underscore. {\em BayesX} is case-sensitive, i.e. #myvar#, #Myvar#
and #MYVAR# are three distinct variable names.

\section{Examples: The credit scoring data set}

This section illustrates some of the
methods described in this chapter using the credit scoring data
set (see \autoref{creditdata} for a description).

We start by illustrating how to code categorical variables
according to either dummy or effect coding. This will be useful in
regression models, where all categorical covariates must be coded
in dummy or effect coding before they can be added to the model.

We start by creating the {\em dataset object} #credit# and proceed
by reading the data using the #infile# command.

#> dataset credit# \\
#> credit.infile using c:\bayesx\examples\credit.raw#

We can now generate new variables to obtain dummy coded versions
of the categorical covariates
#account#, #payment#, #intuse# and #marstat#:

#> credit.generate account1  = 1*(account=1)# \\
#> credit.generate account2  = 1*(account=2)# \\
#> credit.generate payment1 = 1*(payment=1)# \\
#> credit.generate intuse1 = 1*(intuse=1)# \\
#> credit.generate marstat1 = 1*(marstat=1)#

The reference categories are chosen to be 3 for #account# and 2
for the other variables. Alternatively, we could code the
variables according to effect coding. This is achieved with the
following program code:

#> credit.generate account_eff1  = 1*(account=1)-1*(account=3)# \\
#> credit.generate account_eff2  = 1*(account=2)-1*(account=3)# \\
#> credit.generate payment_eff1 = 1*(payment=1)-1*(payment=2)# \\
#> credit.generate intuse_eff1 = 1*(intuse=1)-1*(intuse=2)# \\
#> credit.generate marstat_eff1 = 1*(marstat=1)-1*(marstat=2)#


