\addcontentsline{toc}{chapter}{Bibliography}

\begin{thebibliography}{99}

\label{stepwiseregreferences}

\harvarditem{Albert \& Chib}{1993}{AlbChi93}
 {\scshape Albert, J. \& Chib, S.} (1993).
 Bayesian analysis of binary and polychotomous response data.
 {\it Journal of the American Statistical Association}, {\bf 88}, 669--679.

\harvarditem{Andersson \& Madigan}{1997a}{Andetal97a}
 {\scshape Andersson, S. A., Madigan, D., \& Perlman, M. D.} (1997a).
 A Characterization of Markov equivalence Classes for Acyclic Digraphs.
 {\it The Annals of Statistics}, {\bf 25}, 505--541.

\harvarditem{Andersson, Madigan \& Perlman}{1997b}{Andetal97b}
 {\scshape Andersson, S. A., Madigan, D., \& Perlman, M. D.} (1997b).
 On the Markov equivalence of Chain Graphs, Undirected Graphs, and Acyclic Digraphs.
 { \it Scandinavian Journal of Statistics}, {\bf  24}, 81--102.

\harvarditem{Belitz}{2007}{Bel07}
 {\scshape Belitz, C.} (2007).
 {\it Model Selection in Generalized Structured Additive Regression Models.}
 PhD Thesis, University of Munich.

\harvarditem{Belitz \& Lang}{2008}{BelLan08}
 {\scshape Belitz, C. \& Lang, S.} (2008).
 Simultaneous selection of variables and smoothing parameters in structured additive regression models.
{\it Computational Statistics and Data Analysis}, {\bf 53} , 61-81.

\harvarditem{Besag et al.}{1995}{BesGreHig95}
 {\scshape Besag, J., Green, P., Higdon, D. \& Mengersen, K.} (1995).
 Bayesian computation and stochastic systems (with discussion).
 {\it Statistical Science}, {\bf 10}, 3--66.

\harvarditem{Besag \& Kooperberg}{1995}{BesKoo95}
 {\scshape Besag, J. \& Kooperberg, C.} (1995).
 On conditional and intrinsic autoregressions.
 {\it Biometrika}, {\bf 82}, 733--746.

\harvarditem{Besag, York \& Molli\'{e}}{1991}{BesYorMol91}
 {\scshape Besag, J., York, J. \& Molli\'{e}, A.} (1991).
 Bayesian image restoration with two applications in spatial statistics (with discussion).
 {\it Annals of the Institute of Statistical Mathematics}, {\bf 43}, 1--59.

\harvarditem{Biller}{2000}{Bil00}
 {\scshape Biller, C.} (2000).
 {\it Bayesianische Ans\"{a}tze zur nonparametrischen Regression.}
 Skaker Verlag, Aachen.

\harvarditem{Brezger \& Lang}{2006}{BreLan06}
 {\scshape Brezger, A. \& Lang, S.} (2006).
 Generalized additive regression based on Bayesian P-splines.
 {\it Computational Statistics and Data Analysis} {\bf 50}, 967--991.

\harvarditem{Brezger \& Lang}{2008}{BreLan08}
 {\scshape Brezger, A. \& Lang, S.} (2008).
Simultaneous Probability statements for Bayesian P-splines. {\it
Statistical Modelling}, {\bf 8},
      141--168.

\harvarditem{Brooks}{1998}{Bro98}
 {\scshape Brooks, S. P.} (1998).
 Markov Chain Monte Carlo Method and its Application.
 {\it The Statistician}, {\bf 47}, 69--100.

\harvarditem{Burnham \& Anderson}{1998}{BurAnd98}
 {\scshape Burnham, K. P. \& Anderson, D. R.} (1998).
 {\it Model Selection and Multimodel Inference.}
 Springer, New York.

\harvarditem{Chambers \& Hastie}{1991}{ChaHas91}
 {\scshape Chambers, J. M. \& Hastie, T.} (1991).
 {\it Statistical Models in S.}
 Chapman and Hall.

\harvarditem{Clayton}{1996}{Cla96}
 {\scshape Clayton, D.} (1996).
 Generalized linear mixed models.
 In: Gilks, W., Richardson S. \& Spiegelhalter D. (eds.),
{\it Markov Chain Monte Carlo in Practice}, 275--301.
 London: Chapman and Hall.

\harvarditem{Chen \& Dey}{2000}{CheDey00}
 {\scshape Chen, M. H. \& Dey, D. K.} (2000).
 Bayesian Analysis for Correlated Ordinal Data Models.
 In: Dey, D. K., Ghosh, S. K. \& Mallick, B. K. (eds.),
 {\it Generalized linear models: A Bayesian perspective}, 133--159.
 Marcel Dekker, New York.

\harvarditem{Chib \& Greenberg}{1995}{ChiGre95}
 {\scshape Chib, S. \& Greenberg, E.} (1995).
 Understanding the Metropolis-Hastings Algorithm.
 {\it The American Statistician}, {\bf 49}, 327--335.

\harvarditem{Chickering}{1995}{Chi95}
 {\scshape Chickering, D. M.} (1995).
 A Transformational Characterization of Equivalent Bayesian Network Structures.
 In: Besnard, P. \& Hanks, S. (eds.),
 {\it Uncertainty in Artificial Intelligence, Proceedings of the Eleventh Conference},  87--98.
 San Francisco: Morgan Kaufmann.

\harvarditem{Eilers \& Marx}{1996}{EilMar96}
 {\scshape Eilers, P. H. C. \& Marx, B. D.} (1996).
 Flexible smoothing using B-splines and penalized likelihood (with comments and rejoinder).
 {\it Statistical Science}, {\bf 11}, 89--121.

\harvarditem{Fahrmeir, Kneib \& Lang}{2004}{FahKneLan04}
 {\scshape Fahrmeir, L., Kneib, T. \& Lang, S.} (2004).
 Penalized structured additive regression for space-time data: A Bayesian perspective.
 {\it Statistica Sinica}, {\bf 14}, 715--745.

\harvarditem{Fahrmeir et al.}{2013}{fahkne13}
{\scshape Fahrmeir, L.,  Kneib, T., Lang, S. \& Marx, B.}(2013)
{\it Regression: Models, Methods and Applications.}
 New York: Springer-Verlag.


\harvarditem{Fahrmeir \& Lang}{2001}{FahLan01a}
 {\scshape Fahrmeir, L. \& Lang, S.} (2001a).
 Bayesian Inference for Generalized Additive Mixed Models Based on Markov Random Field Priors.
 {\it Journal of the Royal Statistical Society C}, {\bf 50}, 201--220.

\harvarditem{Fahrmeir \& Lang}{2001}{FahLan01b}
 {\scshape Fahrmeir, L. \& Lang, S.} (2001b).
 Bayesian Semiparametric Regression Analysis of Multicategorical Time-Space Data.
 {\it Annals of the Institute of Statistical Mathematics}, {\bf 53}, 10--30.

\harvarditem{Fahrmeir \& Osuna}{2006}{FahOsu06}
 {\scshape Fahrmeir, L.\& Osuna, L.} (2006).
 Structured additive regression for overdispersed and zero-inflated count data.
 {\it Applied Stochastic Models in Business and Industry}, {\bf 22}, 351--369

 \harvarditem{Fahrmeir \& Tutz}{2001}{FahTut01}
 {\scshape Fahrmeir, L. \& Tutz, G.} (2001).
 {\it Multivariate Statistical Modelling based on Generalized Linear Models.}
 New York: Springer-Verlag.

\harvarditem{Fronk \& Giudici}{2004}{FroGiu00}
 {\scshape Fronk, E.-M. \& Giudici, P.} (2004).
 Markov Chain Monte Carlo model selection for DAG Models.
 {\it Statistical Methods and Applications}, {\bf 13}, 259-273.

\harvarditem{Gamerman}{1997}{Gam97}
 {\scshape Gamerman, D.} (1997).
 Efficient Sampling from the posterior distribution in generalized linear models.
 {\it Statistics and Computing}, {\bf 7}, 57--68.

\harvarditem{Geiger \& Heckerman}{1994}{GeiHec94}
 {\scshape Geiger, D. \& Heckerman, D.} (1994).
 Learning Gaussian Networks.
 {\it Proceedings of the Tenth Conference on Uncertainty in Artificial Intelligence}, 235--243.

\harvarditem{Geiger \& Heckerman}{1999}{GeiHec99}
 {\scshape Geiger, D. \& Heckerman, D.} (1999).
 Parameter priors for directed acyclic graphical models and the characterisation of several probability distributions.
 Submitted for publication.

\harvarditem{Gelfand, Sahu \& Carlin}{1996}{GelSahCar96}
 {\scshape Gelfand, A. E., Sahu, S. K. \& Carlin, B. P.} (1996).
 Efficient Parametrizations for Genera\-lized Linear Mixed Models.
 In: Bernardo, J. M., Berger, J. O., Dawid, A. P. \& Smith, A. F. M. (eds.),
 {\it Bayesian Statistics 5}, 165--180.
 Oxford University Press.

\harvarditem{George \& Liu}{1981}{GeoLiu81}
 {\scshape George, A. \& Liu, J.W.} (1981).
 {\it Computer Solution of Large Sparse Positive Definite Systems.}
 Series in computational mathematics, Prentice-Hall.

\harvarditem{Gilks, Richardson \& Spiegelhalter}{1996}{GilRicSpi96}
 {\scshape Gilks, W. R., Richardson, S., \& Spiegelhalter, D. J.} (1996).
 {\it Markov Chain Monte Carlo in Practice.}
 Chapman and Hall, London.

\harvarditem{Giudici \& Green}{1999}{GiuGre99}
 {\scshape Giudici, P. \& Green, P. J.} (1999).
 Decomposable Graphical Gaussian Model Determination.
 {\it Biometrika}, {\bf 86}, 785--801.

\harvarditem{Green}{1995}{Gre95}
 {\scshape Green, P. J.} (1995).
 Reversible Jump Markov Chain Monte Carlo Computation and Bayesian Model Determination.
 {\it Biometrika}, {\bf  82}, 711--32.

\harvarditem{Green}{2001}{Gre01}
 {\scshape Green, P. J.} (2001).
 A Primer in Markov Chain Monte Carlo.
 In: Barndorff-Nielsen, O. E., Cox, D. R. \& Kl\"{u}ppelberg, C. (eds.),
 {\it Complex Stochastic Systems}, 1--62.
 Chapmann and Hall, London.

\harvarditem{Green \& Silverman}{1994}{GreSil94}
 {\scshape Green, P. J. \& Silverman, B.} (1994).
 {\it Nonparametric Regression and Generalized Linear Models.}
 Chapman and Hall, London.

\harvarditem{Griffin \& Brown}{2005}{GriBro05}
 {\scshape Griffin, J. E., and Brown, P. J.} (2005). Alternative prior distributions
 for variable selection with very many more variables than observations. Technical
 report, University of Warwick, Dept. of Statistics.

\harvarditem{Hastie \& Tibshirani}{1990}{HasTib90}
 {\scshape Hastie, T. \& Tibshirani, R.} (1990).
 {\it Generalized additive models.}
 Chapman and Hall, London.

\harvarditem{Hastie \& Tibshirani}{1993}{HasTib93}
 {\scshape Hastie, T. \& Tibshirani, R.} (1993).
 Varying-coefficient Models.
 {\it Journal of the Royal Statistical Society B}, {\bf 55}, 757--796.

\harvarditem{Hastie \& Tibshirani}{2000}{HasTib00}
 {\scshape Hastie, T. \& Tibshirani, R.} (2000).
 Bayesian Backfitting.
 {\it Statistical Science}, {\bf 15}, 193--223.

\harvarditem{Hastie, Tibshirani \& Firedman}{2001}{HasTibFri01}
 {\scshape Hastie, T., Tisbshirani, R. \& Friedman, J.} (2001).
 {\it The Elements of Statistical Learning: Data Mining, Inference and Prediction.}
 New York: Sprigner-Verlag.

\harvarditem{Hennerfeind, Brezger \& Fahrmeir}{2006}{HenBreFah06}
 {\scshape Hennerfeind, A., Brezger, A. \& Fahrmeir, L.} (2006).
 Geoadditive survival models.
 {\it Journal of the American Statistical Association}, {\bf 101}, 1065--1075.

\harvarditem{Ishwaran \& Rao}{2005}{IshRao05}
 {\scshape Ishwaran, H., and Rao, S. J.} (2005). Spike and Slab Variable
 Selection: Frequentist and Bayesian Strategies. {\it The Annals of Statistics},
 {\bf 33}, 730-773.

\harvarditem{Kandala et al.}{2001}{Kanetal01}
 {\scshape Kandala, N. B., Lang, S., Klasen, S. and Fahrmeir, L.} (2001).
 Semiparametric Analysis of the Socio-Demographic and Spatial Determinants of Undernutrition in Two African Countries.
 {\it Research in Official Statistics}, {\bf 1}, 81--100.

\harvarditem{Klein, Kneib \& Lang}{2014}{kleknelan14a}
{\scshape Klein, N., Kneib, T. \& Lang, S.}(2014).
Bayesian structured additive distributional regression.
{\it Under revision for Annals of Applied Statistics}.


\harvarditem{Kneib}{2006}{Kne06}
 {\scshape Kneib, T.} (2006).
 Geoadditive hazard regression for interval censored survival times.
 {\it Computational Statistics and Data Analysis}, {\bf 51}, 777--792

\harvarditem{Kneib \& Hennerfeind}{2006}{KneHen06}
 {\scshape Kneib, T. \& Hennerfeind, A.} (2006).
 Bayesian Semiparametric Multi-State Models.
 {\it Statistical Modelling}, {\bf 8}, 169--198.

\harvarditem{Kneib \& Fahrmeir}{2006}{KneFah06}
 {\scshape Kneib, T. \& Fahrmeir, L.} (2006).
 Structured additive regression for categorical space-time data: A mixed model approach.
 {\it Biometrics}, {\bf 62}, 109--118.

\harvarditem{Kneib \& Fahrmeir}{2007}{KneFah07}
 {\scshape Kneib, T. \& Fahrmeir, L.} (2007).
 A mixed model approach to structured hazard regression.
 {\it Scandinavian Journal of Statistics}, {\bf 34}, 207--228.

\harvarditem{Kneib et al.}{2009}{Kne09}
 {\scshape Kneib, T., Konrath, S. \& Fahrmeir, L.} (2009). High-dimensional
 Structured Additive Regression Models: Bayesian Regularisation, Smoothing and Predictive
 Performance. Department of Statistics, Technical Report No. 46, LMU Munich.

\harvarditem{Knorr-Held}{1999}{KnoHel99}
 {\scshape Knorr-Held, L.} (1999).
 Conditional Prior Proposals in Dynamic Models.
 {\it Scandinavian Journal of Statistics}, {\bf 26}, 129--144.

\harvarditem{Konrath et al.}{2008}{Kon08}
 {\scshape Konrath, S., Kneib, T., Fahrmeir, L.} (2008). Bayesian Regularisation
 in Structured Additive Regression Models for Survival Data. Department of Statistics,
 Technical Report No.35, LMU Munich.

\harvarditem{Lang \& Brezger}{2004}{LanBre04}
 {\scshape Lang, S. \& Brezger, A.} (2004).
 Bayesian P-splines.
 {\it Journal of Computational and Graphical Statistics}, {\bf 13}, 183--212.

\harvarditem{Lang et al.}{2014}{LanUml14}
 {\scshape Lang, S., Umlauf, N., Wechselberger, P., Harttgen, K.  \& Kneib, T.} (2014).
 Multilevel Structured Additive Regression.
 {\it Statistics and Computing}, {\bf 24}, 223--238.

\harvarditem{Lauritzen}{1996}{Lau96}
 {\scshape Lauritzen, S. L.} (1996).
 {\it Graphical Models},
 Clarendon Press, Oxford.

\harvarditem{Lin \& Zhang}{1999}{LinZha99}
 {\scshape Lin, X. \& Zhang, D.} (1999).
 Inference in generalized additive mixed models by using smoothing splines.
 {\it Journal of the Royal Statistical Society B}, {\bf 61}, 381--400.

\harvarditem{McCullagh \& Nelder}{1989}{McCNel89}
 {\scshape McCullagh, P. \& Nelder, J. A.} (1989).
 {\it Generalized Linear Models.}
 Chapman and Hall, London.

\harvarditem{Osuna}{2004}{Osu04}
 {\scshape Osuna, L.} (2004)
 {\it Semiparametric Bayesian Count Data Models}.
 Dr. Hut Verlag, M\"{u}nchen.

\harvarditem{Park \& Casella}{2008}{ParCas08}
 {\scshape Park, T., and Casella, G.} (2008). The Bayesian Lasso. {\it Journal
 of the American Statistical Association}, {\bf 482}, 681-686.

\harvarditem{Rue}{2001}{Rue01}
 {\scshape Rue, H.} (2001).
 Fast Sampling of Gaussian Markov Random Fields with Applications.
 {\it Journal of the Royal Statistical Society B}, {\bf 63}, 325--338.

\harvarditem{Tierney}{1983}{Tie83}
 {\scshape Tierney, L.} (1983).
A space-efficient recursive
procedure for estimating a quantile of an
unknown distribution. {\it SIAM Journal on
Scientific and Statistical Computing}, {\bf 4},
706–-11.

\harvarditem{Spiegelhalter et al.}{2002}{SpiBesCar02}
 {\scshape Spiegelhalter, D. J., Best, N. G., Carlin, B. P. \& van der Linde, A.} (2002).
 Bayesian measures of model complexity and fit.
 {\it Journal of the Royal Statistical Society B}, {\bf 65}, 583--639.

\harvarditem{Wood}{2006a}{Woo06a}
 {\scshape Wood, S. N.} (2006a).
 {\it Generalized Additive Models: An Introduction with R.}
 Chapman and Hall.

\harvarditem{Wood}{2006b}{Woo06b}
 {\scshape Wood, S. N.} (2006b).
 On confidence intervals for GAMs based on penalized regression splines.
 {\it Australian and New Zealand Journal of Statistics}, {\bf 48}, 445--464.
\end{thebibliography}
