\documentclass[11pt,a4paper,twoside]{bayesxreport}

\input{manual_preliminaries.tex}

 \renewcommand{\chaptername}{Tutorial}

 \externaldocument{manual}
 \externaldocument{manual_star}

\makeindex

\begin{document}

\MakeShortVerb{\|}

\preface{2.00}{Tutorials}

\newpage

\chapter[A tutorial
on Bayesian semiparametric regression using MCMC
techniques]{Determinants of childhood undernutrition in Zambia: A
tutorial on Bayesian semiparametric regression using MCMC
techniques} \label{zambiaanalysis}

\input{mcmc.tex}

\chapter[A tutorial
on Bayesian semiparametric regression using GLMM
methodology]{Determinants of childhood undernutrition in Zambia: A
tutorial on Bayesian semiparametric regression using GLMM
methodology} \label{remlregzambiaanalysis}

\input{reml.tex}

\chapter[Simultaneously selecting variables and smoothing parameters] {Determinants of childhood undernutrition in Zambia: A
tutorial on a method for simultaneously selecting variables and
smoothing parameters in semiparametric regression models}
\label{zambia_step_analysis}

\input{step.tex}


\addcontentsline{toc}{chapter}{References}

\begin{thebibliography}{99}

\harvarditem{Belitz \& Lang}{2008}{bellan08} Belitz, C. \& Lang, S.  (2001): Simultaneous selection of variables and smoothing
parameters in structured additive regression models. {\it Computational Statistics and Data Analysis}, {\bf 53}, 61--81.

\harvarditem{Brezger \& Lang}{2006}{brelan05} Brezger, A. \& Lang, S., 2006: Generalized structured additive regression based
on Bayesian P-splines. {\it Computational Statistics and Data Analysis}, {\bf 50}, 967--991.

\harvarditem{Fahrmeir, Kneib \& Lang}{2004}{fahkne04} Fahrmeir, L., Kneib, T. \& Lang, S., 2004: Penalized structured additive
regression for space-time data: A Bayesian perspective, {\it Statistica Sinica}, {\bf 14}, 731--761 .

\harvarditem{Fahrmeir \& Lang}{2001a}{fahlan01a} Fahrmeir, L. \& Lang, S., 2001a: Bayesian Inference for Generalized Additive
Mixed Models Based on Markov Random Field Priors. {\it Journal of the Royal Statistical Society C}, {\bf 50}, 201--220.

\harvarditem{Fahrmeir \& Lang}{2001b}{fahlan01b} Fahrmeir, L. \& Lang, S., 2001: Bayesian Semiparametric Regression Analysis of
Multicategorical Time-Space Data. {\it Annals of the Institute of Statistical Mathematics}, {\bf 53}, 10--30

\harvarditem{Fahrmeir \& Osuna}{2006}{fahosu03} Fahrmeir, L. \& Osuna, L. (2003), Structured additive regression for
overdispersed and zero-inflated count data. {\it Applied Stochastic Models in Business and Industry}, {\bf 22}, 351--369.

\harvarditem{George \& Liu}{1981}{geoliu81} George, A. \& Liu, J.W. 1981: {\it Computer Solution of Large Sparse Positive
Definite Systems}, Prentice-Hall.

\harvarditem{Hennerfeind, Brezger \& Fahrmeir}{2006}{henfah03} Hennerfeind, A., Brezger, A. \& Fahrmeir, L., 2003: Geoadditive
survival models. {\it Journal of the American Statistical Association}, {\bf 101}, 1065--1075

\harvarditem{Kandala et~al.}{2001}{kanlan01} Kandala, N. B., Lang, S., Klasen, S. \& Fahrmeir, L. (2001): Semiparametric
Analysis of the Socio-Demographic and Spatial Determinants of Undernutrition in Two African Countries. {\it Research in
Official Statistics}, {\bf 1}, 81--100.

 \harvarditem{Kneib}{2006}{kneib2006b} {Kneib, T.} (2006) Geoadditive hazard regression for interval censored
 survival times. {\it Computational Statistics and Data Analysis}, {\bf 51}, 777--792.

 \harvarditem{Kneib \& Fahrmeir}{2006}{kneib2006a} {Kneib, T. \& Fahrmeir, L.} (2006) Structured additive
 regression for categorical space-time data: A mixed model approach. {\it Biometrics}, {\bf 62}, 109--118.

 \harvarditem{Kneib \& Fahrmeir}{2007}{kneib2007} {Kneib, T. \& Fahrmeir, L.} (2007) A mixed model approach for
 geoadditive hazard regression. {\it Scandinavian Journal of Statistics}, {\bf 34} 207-228.

\harvarditem{Lang \& Brezger}{2004}{lanbre04} Lang, S. \& Brezger, A., 2004: Bayesian P-splines. {\it Journal of Computational
and Graphical Statistics}, {\bf 13}, 183--212.

\harvarditem{Spiegelhalter et~al.}{2002}{spibes02} Spiegelhalter, D.J., Best, N.G., Carlin, B.P. \& van der Linde, A. (2002):
Bayesian measures of model complexity and fit. {\it Journal of the Royal Statistical Society B}, {\bf 65}, 583--639

\end{thebibliography}

\end{document}
