\documentclass[a4paper]{article}

\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage[dvips]{epsfig}
\usepackage[dcu]{harvard}
\usepackage{dcolumn}
\usepackage{shortvrb}
\usepackage{parskip}
\usepackage{multicol}

\sloppy
\parindent0em
\topmargin-1.5cm \textheight23cm \textwidth15cm
\oddsidemargin0.5cm

\MakeShortVerb{\|}

\begin{document}
\title{Bayesian Semiparametric Regression Based on MCMC Simulation Techniques: A Tutorial}
\author{Thomas Kneib, Stefan Lang and Andreas Brezger\\ [.25cm]
\normalsize Department of Statistics, University of Munich. }
\maketitle

\begin{abstract}
This tutorial demonstrates the usage of {\it BayesX} for analysing Bayesian semiparametric regression models based on MCMC
simulation techniques. As an example, we consider data on undernutrition of children in Zambia. The tutorial is designed to be
self-contained and describes all features of {\it BayesX} in detail, that will be needed throughout the tutorial. Therefore it
may also serve as a first introduction into the general usage of {\it BayesX}. Note that the tutorial is written for the
version of {\it BayesX} including a graphical user interface and graphics facilities. While all regression-related commands can
be executed in the command line version as well, the graphics facilities will not be available.
\end{abstract}

\tableofcontents

\newpage

\input{mcmc.tex}

\addcontentsline{toc}{section}{References}

\begin{thebibliography}{99}

\harvarditem{Brezger \& Lang}{2006}{brelan05} Brezger, A. \& Lang,
S., 2006: Generalized structured additive regression based on
Bayesian P-splines. {\it Computational Statistics and Data
Analysis}, {\bf 50}, 967--991.

\harvarditem{Fahrmeir, Kneib \& Lang}{2004}{fahkne04} Fahrmeir, L., Kneib,
T. \& Lang, S., 2004: Penalized structured additive regression for
space-time data: A Bayesian perspective, {\it Statistica Sinica},
{\bf 14}, 731--761 .

\harvarditem{Fahrmeir \& Lang}{2001a}{fahlan01a} Fahrmeir, L. \&
Lang, S., 2001a: Bayesian Inference for Generalized Additive Mixed
Models Based on Markov Random Field Priors. {\it Journal of the
Royal Statistical Society C}, {\bf 50}, 201--220.

\harvarditem{Fahrmeir \& Lang}{2001b}{fahlan01b} Fahrmeir, L. \&
Lang, S., 2001: Bayesian Semiparametric Regression Analysis of
Multicategorical Time-Space Data. {\it Annals of the Institute of
Statistical Mathematics}, {\bf 53}, 10--30

\harvarditem{Fahrmeir \& Osuna}{2006}{fahosu03} Fahrmeir, L. \&
Osuna, L. (2003), Structured additive regression for overdispersed and zero-inflated count data.
{\it Applied Stochastic Models in Business and Industry}, {\bf 22}, 351--369.

\harvarditem{George \& Liu}{1981}{geoliu81} George, A. \& Liu,
J.W. 1981: {\it Computer Solution of Large Sparse Positive
Definite Systems}, Prentice-Hall.

\harvarditem{Hennerfeind, Brezger \& Fahrmeir}{2006}{henfah03} Hennerfeind, A.,
Brezger, A. \& Fahrmeir, L., 2003: Geoadditive survival models.
{\it Journal of the American Statistical Association}, {\bf 101}, 1065--1075

\harvarditem{Kandala et~al.}{2001}{kanlan01} Kandala, N. B., Lang,
S., Klasen, S. \& Fahrmeir, L. (2001): Semiparametric Analysis of
the Socio-Demographic and Spatial Determinants of Undernutrition
in Two African Countries. {\it Research in Official Statistics},
{\bf 1}, 81--100.

\harvarditem{Lang \& Brezger}{2004}{lanbre04} Lang, S. \& Brezger,
A., 2004: Bayesian P-splines. {\it Journal of Computational and
Graphical Statistics}, {\bf 13}, 183--212.

\harvarditem{Spiegelhalter et~al.}{2002}{spibes02} Spiegelhalter,
D.J., Best, N.G., Carlin, B.P. \& van der Linde, A. (2002):
Bayesian measures of model complexity and fit. {\it Journal of the
Royal Statistical Society B}, {\bf 65}, 583--639

\end{thebibliography}


\end{document}
