\documentclass[a4paper]{article}

\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage[dvips]{epsfig}
\usepackage{dcolumn}
\usepackage{shortvrb}
\usepackage{parskip}
\usepackage{multicol}
\usepackage[dcu]{harvard}

\sloppy
\parindent0em
\topmargin-1.5cm \textheight23cm \textwidth15cm
\oddsidemargin0.5cm

\MakeShortVerb{\|}

\begin{document}

\title{Bayesian Semiparametric Regression Based on Mixed Model Methodology: A Tutorial}
\author{Thomas Kneib, Stefan Lang and Andreas Brezger\\ [.25cm]
\normalsize Department of Statistics, University of Munich. } \maketitle

\begin{abstract}
This tutorial demonstrates the usage of {\it BayesX} for analysing Bayesian semiparametric regression models based on mixed
model methodology. As an example, we consider data on undernutrition of children in Zambia. The tutorial is designed to be
self-contained and describes all features of {\it BayesX} in detail, that will be needed throughout the tutorial. Therefore it
may also serve as a first introduction into the general usage of {\it BayesX}. Note that the tutorial is written for the
version of {\it BayesX} including a graphical user interface and graphics facilities. While all regression-related commands can
be executed in the command line version as well, the graphics facilities will not be available.
\end{abstract}

\tableofcontents

\newpage

\input{reml.tex}

\addcontentsline{toc}{section}{References}

\begin{thebibliography}{99}
\harvarditem{Fahrmeir, Kneib \& Lang}{2004}{fahkne04} Fahrmeir, L., Kneib, T. \& Lang, S., 2004: Penalized structured additive
regression for space-time data: A Bayesian perspective, {\it Statistica Sinica}, {\bf 14}, 731--761 .

\harvarditem{Fahrmeir \& Lang}{2001}{fahlan01b} Fahrmeir, L. \& Lang, S., 2001: Bayesian Semiparametric Regression Analysis of
Multicategorical Time-Space Data. {\it Annals of the Institute of Statistical Mathematics}, {\bf 53}, 10--30

\harvarditem{Kandala et~al.}{2001}{kanlan01} Kandala, N. B., Lang, S., Klasen, S. \& Fahrmeir, L. (2001): Semiparametric
Analysis of the Socio-Demographic and Spatial Determinants of Undernutrition in Two African Countries. {\it Research in
Official Statistics}, {\bf 1}, 81--100.

 \harvarditem{Kneib}{2006}{kneib2006b} {Kneib, T.} (2006) Geoadditive hazard regression for interval censored
 survival times. {\it Computational Statistics and Data Analysis}, {\bf 51}, 777--792.

 \harvarditem{Kneib \& Fahrmeir}{2006}{kneib2006a} {Kneib, T. \& Fahrmeir, L.} (2006) Structured additive
 regression for categorical space-time data: A mixed model approach. {\it Biometrics}, {\bf 62}, 109--118.

 \harvarditem{Kneib \& Fahrmeir}{2007}{kneib2007} {Kneib, T. \& Fahrmeir, L.} (2007) A mixed model approach for
 geoadditive hazard regression. {\it Scandinavian Journal of Statistics}, {\bf 34} 207-228.


\end{thebibliography}


\end{document}
