\documentclass[a4paper]{article}

\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage[dvips]{epsfig}
\usepackage{dcolumn}
\usepackage{shortvrb}
\usepackage{parskip}
\usepackage{multicol}
\usepackage[dcu]{harvard}

\sloppy
\parindent0em
\topmargin-1.5cm \textheight23cm \textwidth15cm
\oddsidemargin0.5cm

\MakeShortVerb{\|}

\begin{document}

\title{Simultaneously Selecting Variables and Smoothing Parameters in Semiparametric Regression Models: A Tutorial}
\author{Christiane Belitz and Stefan Lang\\ [.25cm]
\normalsize Department of Statistics, University of Munich. } \maketitle

\begin{abstract}
This tutorial demonstrates the usage of {\it BayesX} for simultaneously selecting variables and smoothing parameters in
semiparametric regression models. As an example, we consider data on undernutrition of children in Zambia. The tutorial is
designed to be self-contained and describes all features of {\it BayesX} in detail, that will be needed throughout the
tutorial. Therefore it may also serve as a first introduction into the general usage of {\it BayesX}. Note that the tutorial is
written for the version of {\it BayesX} including a graphical user interface and graphics facilities. While all
regression-related commands can be executed in the command line version as well, the graphics facilities will not be available.
\end{abstract}

\tableofcontents

\newpage

\input{step.tex}

\addcontentsline{toc}{section}{References}

\begin{thebibliography}{99}

\harvarditem{Belitz \& Lang}{2008}{bellan08} Belitz, C. \& Lang, S.  (2001): Simultaneous selection of variables and smoothing
parameters in structured additive regression models. {\it Computational Statistics and Data Analysis}, {\bf 53}, 61--81.

\harvarditem{George \& Liu}{1981}{geoliu81} George, A. \& Liu, J.W. 1981: {\it Computer Solution of Large Sparse Positive
Definite Systems}, Prentice-Hall.

\harvarditem{Kandala et~al.}{2001}{kanlan01} Kandala, N. B., Lang, S., Klasen, S. \& Fahrmeir, L. (2001): Semiparametric
Analysis of the Socio-Demographic and Spatial Determinants of Undernutrition in Two African Countries. {\it Research in
Official Statistics}, {\bf 1}, 81--100.


\end{thebibliography}

\end{document}
